---
title: "Visualizing commute flows in Wisconsin"
description: |
  LODES provides origin-destination data for connecting work and home. This article looks at patters in those flows in Wisconsin, and especially in Madison
author:
  - name: Nora Jones 
    url: https://example.com/norajones
    affiliation: Spacely Sprockets
    affiliation_url: https://example.com/spacelysprokets
date: "`r Sys.Date()`"
output: distill::distill_article
---


```{r}
library(tidyverse)
library(tigris)
library(sf)
library(mapdeck)
```

# Get data

LODES OD files from 2020. Warning. File is about 15 MB!

```{r eval=F}
download.file("https://lehd.ces.census.gov/data/lodes/LODES7/wi/od/wi_od_main_JT00_2018.csv.gz", "data/wi_od_main_JT00_2018.csv.gz")
```

Saving data as RDS file for faster processing, then deleting the original file.

```{r eval=F}

OD <- read_csv("data/wi_od_main_JT00_2018.csv.gz")
saveRDS(OD, file = "data/wi_od_main_JT00_2018.rds")
file.remove("data/wi_od_main_JT00_2018.csv.gz")
```

The file structure is explained [here](https://lehd.ces.census.gov/data/lodes/LODES7/LODESTechDoc7.5.pdf), page 6. There are two geography identifier variables, `w_geocode` and `h_geocode`, and a number of variables for the number of jobs for that geography pair. `S000` is the one for total jobs.

```{r}
OD <- readRDS("data/wi_od_main_JT00_2018.rds")

OD <- OD %>% 
  mutate(across(1:2, as.character)) %>%  #change GEOID columns to character to be safe
  select(w_geocode, h_geocode, S000) #keep only home/work, total jobs
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


The [`mapdeck` arc visualization](https://symbolixau.github.io/mapdeck/articles/layers.html) requires either two pairs of coordinates or `sfc` columns.

I will get TIGER geographies with the `tigris` package, calculate centroids with `sf`, and then merge with the LODES data. Note that LODES uses the 2010 vintage for its geographies, whereas `tigris` defaults to the most recent vintage.

```{r fetch-WI-blocks}
library(tigris)
options(tigris_use_cache = TRUE)

wi <- blocks("WI", year = 2010)
```

```{r calculate-centroids}
library(sf)
wi_centroid <- wi %>% 
  select(GEOID10, geometry) %>% 
  st_centroid()
```

Let's look at the distribution of the number of jobs per origin-destination pair:

```{r}
OD %>% 
  ggplot(aes(S000)) +
  geom_density()
```

This does not look terribly promising -- maybe block pairs are the wrong unit of analysis, with a huge number of them only having a single job-home pair. Aggregation to block groups or tracts may make more sense, but for now I'll proceed with the blocks.

# Commute flows in Wisconsin

The number of block pairs is large, and so I'll sample the 100 blocks with the largest number of commutes.

```{r top-hundred}
OD_100 <- OD %>% 
  arrange(desc(S000)) %>% 
  head(n = 100)
```

Now we can do a double join with the block centroids: First, join on the work geocode, then do another join on the home geocode. This will create a new dataframe with two geometry columns. In order to work with the `mapdeck` package, the dataframe needs to be turned into an `sf` object. Finally, we add a thickness helper variable based on the number of jobs in the OD pair. This variable will specify the width of the arcs in the visualization. 

```{r join-od100}
OD_100 <- OD_100 %>% 
  left_join(wi_centroid, by = c("w_geocode" = "GEOID10")) %>% 
  inner_join(wi_centroid, by = c("h_geocode" = "GEOID10")) %>% 
  mutate(thick = S000/10) %>% 
  rename(origin = geometry.y,
         destination = geometry.x)

OD_100 <- OD_100 %>% 
  st_as_sf()
```

Census blocks can be oddly shaped, and so it's a good idea to plot the actual blocks and not just their centroids.

```{r}
target_blocks_OD_100 <-  wi %>%
  filter(GEOID10 %in% OD_100$w_geocode | GEOID10 %in% OD_100$h_geocode)
```

## Visualizing the flows

Now we can visualize the top-100 commute flows in the state:

```{r}
library(mapdeck)

# function for plotting the map

OD_arcs <- function(OD){
  #create target blocks
  target_blocks <-  wi %>%
  filter(GEOID10 %in% OD$w_geocode | GEOID10 %in% OD$h_geocode)
  #draw map
  
mapdeck(style = mapdeck_style("light"), pitch = 45 ) %>%
  add_arc(
    data = OD
    , layer_id = "arc_layer"
    , origin = "origin"
    , destination = "destination"
    , stroke_from = "#404040"
    , stroke_to = "#ca0020"
    , stroke_width = "thick"
  ) %>% 
  add_polygon(
    data = target_blocks,
    layer_id = "blocks",
    fill_colour = "#54135296",
    fill_opacity = .2,
    stroke_colour = "#404040",
    tooltip = "w_geocode"
  )
}

OD_arcs(OD_100)
```


Wow, it is immediately obvious how dominant commute flows in the Madison area are! And by zooming in you can identify some other patterns:

- In Madison two employers capture almost all commute flows: The UW-Madison downtown^[For some reason, the university reports all(?) of its employees as working at a single location. In reality, UW employment is more spread out.] and electronic health record company Epic in suburban Verona. The only other employer making an appearance is the university hospital (which is a separate entity from the university)
- In Milwaukee, by far the state's largest city, there are only three OD pairs in total: Two point at the [32-story Northwestern Mutual Tower](https://en.wikipedia.org/wiki/Northwestern_Mutual_Tower_and_Commons) downtown; the third leads to the large suburban medical campus that includes the Medical College of Wisconsin, Froedtert Hospital, and the Childrens' Hospital of Wisconsin.
- Sheboygan is probably best known for plumbing product company Kohler, and the Kohler Co. headquarters make a prominent appearance on the map, with 8 arcs pointing at it.
- For the remainder of OD pairs around the state, it's primarily meat processing, healthcare facilities, and tribal facilities that appear as destinations.
- For two OD pairs, home and work are actually in the same block: One is the UW in Platteville (presumably students living and working on campus), and a strangely empty looking wooded block near Hayward. My best guess is that it's a farm or forestry business.


# Madison flows

Given the visual prominence of Madison on the map (and my personal interest in Madison), it makes sense to do a separate analysis just for flows within Dane County.

```{r filter-dane}
# filter to only block with origin and destination within Dane County
# filtering through spatial operations is computationally expensive
# instead filtering by the GEOID: 55 is the state, 025 Dane County

OD_dane <- OD %>% 
  filter(str_detect(w_geocode, "55025") & str_detect(h_geocode, "55025"))


OD_dane <- OD_dane %>% 
  left_join(wi_centroid, by = c("w_geocode" = "GEOID10")) %>% 
  inner_join(wi_centroid, by = c("h_geocode" = "GEOID10")) %>% 
  mutate(thick = S000/10) %>% 
  rename(origin = geometry.y,
         destination = geometry.x)

OD_dane <- OD_dane %>% 
  st_as_sf()
```

```{r}
OD_dane_100 <- OD_dane %>% 
  filter(S000>1) %>% #speeds things up
  arrange(desc(S000)) %>% 
  head(n = 100)

OD_arcs(OD_dane_100)
```

Well, this doesn't look all that different, does it? More arcs, but all of them still point at the UW, Epic, and the university hospital. This doesn't change even if you look at the top-150 OD pairs. Only when you bump it up to 300, a few more work locations appear.

```{r}
OD_dane_300 <- OD_dane %>% 
  arrange(desc(S000)) %>% 
  head(300)

unique(OD_dane_300$w_geocode)
```

Plotting 300 arcs of course is a bit of a mess:

```{r}
OD_arcs(OD_dane_300)
```
If you scroll around enough you see the City-County Building, American Family Insurance, and a block with several employers near the Am Fam campus.

