---
title: "Untitled"
description: |
  A new article created using the Distill format.
author:
  - name: Nora Jones 
    url: https://example.com/norajones
    affiliation: Spacely Sprockets
    affiliation_url: https://example.com/spacelysprokets
date: "`r Sys.Date()`"
output: distill::distill_article
---

# Get data

LODES OD files from 2020. Warning. File is about 15 MB!

```{r eval=F}
download.file("https://lehd.ces.census.gov/data/lodes/LODES7/wi/od/wi_od_main_JT00_2018.csv.gz", "data/wi_od_main_JT00_2018.csv.gz")
```

Saving data as RDS file for faster processing, then deleting the original file.
```{r eval=F}
library(tidyverse)
OD <- read_csv("data/wi_od_main_JT00_2018.csv.gz")
saveRDS(OD, file = "data/wi_od_main_JT00_2018.rds")
file.remove("data/wi_od_main_JT00_2018.csv.gz")
```

The file structure is explained [here](https://lehd.ces.census.gov/data/lodes/LODES7/LODESTechDoc7.5.pdf), page 6. There are two geography identifier variables, `w_geocode` and `h_geocode`, and a number of variables for the number of jobs for that geography pair. `S000` is the one for total jobs.

```{r}
OD <- readRDS("data/wi_od_main_JT00_2018.rds")

OD <- OD %>% 
  mutate(across(1:2, as.character)) %>%  #change GEOID columns to character to be safe
  select(w_geocode, h_geocode, S000) #keep only home/work, total jobs
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r}
url <- 'https://raw.githubusercontent.com/plotly/datasets/master/2011_february_aa_flight_paths.csv'
flights <- read.csv(url)
flights$id <- seq_len(nrow(flights))
flights$stroke <- sample(1:3, size = nrow(flights), replace = T)
flights$start_elev <- sample(100000:1000000, size = nrow(flights), replace = TRUE )

sf_flights <- sfheaders::sf_point(flights, x = "start_lon", y = "start_lat", z = "start_elev", keep = TRUE)

destination <- sfheaders::sfc_point(flights, x = "end_lon", y = "end_lat", z = "start_elev")

sf_flights$destination <- destination

library(mapdeck)
mapdeck(
) %>%
 add_arc(
   data = sf_flights
   , origin = 'geometry'
   , destination = 'destination'
   , layer_id = 'arcs'
   , stroke_from = "airport1"
   , stroke_to = "airport2"
)
```


The [`mapdeck` arc visualization](https://symbolixau.github.io/mapdeck/articles/layers.html) requires either two pairs of coordinates or `sfc` columns. 

I will get TIGER geographies with the `tigris` package, calculate centroids with `sf`, and then merge with the LODES data.

```{r}
library(tigris)
options(tigris_use_cache = TRUE)

wi <- blocks("WI")
```


```{r}
library(sf)
wi_centroid <- wi %>% 
  select(GEOID10, geometry) %>% 
  st_centroid()
```

Let's look at the distribution of the number of jobs per origin-destination pair:

```{r}
OD %>% 
  ggplot(aes(S000)) +
  geom_density()
```
This does not look terribly promising -- maybe block pairs are the wrong unit of analysis, with a huge number of them only having a single job. Aggregation to block groups or tracts may make more sense, but for now I'll proceed with the blocks.

```{r}
OD_100 <- OD %>% 
  arrange(desc(S000)) %>% 
  head(n = 100)
```

```{r}
OD_100 <- OD_100 %>% 
  left_join(wi_centroid, by = c("w_geocode" = "GEOID10"))

OD_101 <- wi_centroid %>% 
  inner_join(OD_100, by = c("GEOID10" = "w_geocode"))

destination <- wi_centroid %>% 
  inner_join(OD_100, by = c("GEOID10" = "h_geocode"))

OD_101$destination <- destination$geometry

OD_101 <- OD_101 %>% 
  mutate(thick = S000/10)
```


```{r}
library(mapdeck)


mapdeck(style = mapdeck_style("light"), pitch = 45 ) %>%
  add_arc(
    data = OD_101
    , layer_id = "arc_layer"
    , origin = "geometry"
    , destination = "destination"
    , stroke_from = "#404040"
    , stroke_to = "#ca0020"
    , stroke_width = "thick"
  )
```


This is quite surprising to me. I would have expected that the dominant commute pattern would be within cities or metro areas. Instead what we're seeing are a lot of long-distance commute pairs. A lot of them center on Madison (mostly as a destination, but also as an origin), but there also odd pairs of commutes between Platteville and Barron (home of a large turkey processing plant) and others.

Let's see if this changes if instead of all jobs we only look at primary jobs.

```{r}
download.file("https://lehd.ces.census.gov/data/lodes/LODES7/wi/od/wi_od_main_JT00_2018.csv.gz", "data/wi_od_main_JT01_2018.csv.gz")
```

```{r}
OD_primary <- read_csv("data/wi_od_main_JT01_2018.csv.gz")
saveRDS(OD_primary, file = "data/wi_od_main_JT01_2018.rds")
file.remove("data/wi_od_main_JT01_2018.csv.gz")
```
```{r}
OD_primary <- readRDS("data/wi_od_main_JT00_2018.rds")

OD_primary <- OD %>% 
  mutate(across(1:2, as.character)) %>%  #change GEOID columns to character to be safe
  select(w_geocode, h_geocode, S000) #keep only home/work, total jobs
```


```{r}
OD_primary_100 <- OD_primary %>% 
  arrange(desc(S000)) %>% 
  head(n = 100)

destination <- wi_centroid %>% 
  inner_join(OD_primary_100, by = c("GEOID10" = "h_geocode"))

OD_primary_100 <- wi_centroid %>% 
  inner_join(OD_primary_100, by = c("GEOID10" = "w_geocode"))



OD_primary_100$destination <- destination$geometry

OD_primary_100 <- OD_primary_100 %>% 
  mutate(thick = S000/10)

```

```{r}
library(mapdeck)


mapdeck(style = mapdeck_style("light"), pitch = 45 ) %>%
  add_arc(
    data = OD_primary_100
    , layer_id = "arc_layer"
    , origin = "geometry"
    , destination = "destination"
    , stroke_from = "#404040"
    , stroke_to = "#ca0020"
    , stroke_width = "thick"
  )
```

## Get LODES employment data
# file downloaded from https://lehd.ces.census.gov/data/
# WI -> state
# WAC -> workplace geography
# S000 -> segment of the workforce: all jobs
# JT00 -> job type: all jobs
# documentation here: https://lehd.ces.census.gov/data/lodes/LODES7/LODESTechDoc7.4.pdf
jobs <- read_csv("data/lodes/wi_wac_S000_JT00_2017.csv")

jobs <- jobs %>% 
  mutate(GEOID = as.character(w_geocode))

#get TIGER shapefiles for blocks
blocks <- blocks("WI", county = "Dane", year = 2017)
  

# join jobs and block groups
jobs_blocks <- blocks %>% 
  left_join(jobs, by = c("GEOID10" = "GEOID"))

tmap_mode("view")
tm_shape(jobs_blocks) +
  tm_polygons("C000")



